---
title: "Nonlinear Models (ISL 7)"
author: "Dr. Jin Zhou @ UCLA"
date: "May 30, 2023"
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4  
subtitle: Biostat 200C
---


Credit: This note heavily uses material from the books [_An Introduction to Statistical Learning: with Applications in R_](https://www.statlearning.com/) (ISL2) and [_Elements of Statistical Learning: Data Mining, Inference, and Prediction_](https://hastie.su.domains/ElemStatLearn/) (ESL2).

Display system information for reproducibility.

```{r}
sessionInfo()
```


## Overview

- The truth is never linear! Or almost never! 

    But often the linearity assumption is good enough.  
    
- When it's not ...

    - polynomials  
    - step functions  
    - spline  
    - local regression, and
    - generalized additive models  
    
    offer a lot of flexibility, without losing the ease and interpretability of linear models. 

- `wage` vs `age`:



```{r}
#| message: false
library(gtsummary)
library(ISLR2)
library(tidyverse)

# Convert to tibble
Wage <- as_tibble(Wage) %>% print(width = Inf)

# Summary statistics
Wage %>% tbl_summary()

# Plot wage ~ age, GAM fit is display when n >1000
Wage %>%
  ggplot(mapping = aes(x = age, y = wage)) + 
  geom_point() + 
  geom_smooth() +
  labs(title = "Wage changes nonlinearly with age",
       x = "Age",
       y = "Wage (k$)")
```


    
## Polynomial regression

$$
y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \cdots + \beta_d x_i^d + \epsilon_i.
$$



```{r}
# Plot wage ~ age, display order-4 polynomial fit
Wage %>%
  ggplot(mapping = aes(x = age, y = wage)) + 
  geom_point() + 
  geom_smooth(
    method = "lm",
    formula = y ~ poly(x, degree = 4)
    ) +
  labs(
    title = "Degree-4 Polynomial",
    x = "Age",
    y = "Wage (k$)"
    )
```


- Create new variables $X_1 = X$, $X_2 = X^2$, ..., and then treat as multiple linear regression.

- Not really interested in the coefficients; more interested in
the fitted function values at any value $x_0$:
$$
\hat f(x_0) = \hat{\beta}_0 + \hat{\beta}_1 x_0 + \hat{\beta}_2 x_0^2 + \hat{\beta}_3 x_0^3 + \hat{\beta}_4 x_0^4.
$$



```{r}
# poly(age, 4) constructs orthogonal polynomial of degree 1 to degree, all orthogonal to the constant
lmod <- lm(wage ~ poly(age, degree = 4), data = Wage)
summary(lmod)

# poly(age, 4, raw = TRUE) procudes raw othogonal polynomial, which match Python
lmod <- lm(wage ~ poly(age, degree = 4, raw = TRUE), data = Wage)
summary(lmod)
```

:::

- Since $\hat f(x_0)$ is a linear function of the $\hat{\beta}_j$, we can get a simple expression for **pointwise-variances** $\operatorname{Var}[\hat f(x_0)]$ at any value $x_0$. 

- We either fix the degree $d$ at some reasonably low value, or use cross-validation to choose $d$. 

- Can do separately on several variables. Just stack the variables into one matrix, and separate out the pieces afterwards (see GAMs later).

- Polynomial modeling can be done for generalized linear models (logistic regression, Poisson regression, etc) as well. 

- **Caveat**: polynomials have notorious tail behavior. Very bad for extrapolation.


```{r}
#| code-fold: true
library(splines)

# Plot wage ~ age
Wage %>%
  ggplot(mapping = aes(x = age, y = wage)) + 
  geom_point(alpha = 0.25) + 
  # Polynomial regression with degree 14
  geom_smooth(
    method = "lm",
    formula = y ~ poly(x, degree = 14),
    color = "blue"
    ) +
  # Natural cubic spline
  geom_smooth(
    method = "lm",
    formula = y ~ ns(x, df = 14),
    color = "red"
    ) +  
  labs(
    title = "Natural cubic spline (red) vs polynomial regression (blue)",
    subtitle = "Both have df=15",
    x = "Age",
    y = "Wage (k$)"
    )
```


## Piecewise polynomials (regression splines)

- Instead of a single polynomial in $X$ over its whole domain, we can rather use different polynomials in regions defined by **knots**. E.g., a piecewise cubic polynomial with a single knot at $c$ takes the form
$$
y_i = \begin{cases}
\beta_{01} + \beta_{11} x_i + \beta_{21} x_i^2 + \beta_{31} x_i^3 + \epsilon_i & \text{if } x_i < c \\
\beta_{02} + \beta_{12} x_i + \beta_{22} x_i^2 + \beta_{32} x_i^3 + \epsilon_i & \text{if } x_i \ge c
\end{cases}.
$$

- Better to add constraints to the polynomials, e.g., continuity.

- Splines have the "maximum" amount of continuity.

<p align="center">
![](ISL_fig_7_3.pdf){width=600px height=600px}
</p>

### Linear spline

- A **linear spline** with knots at $\xi_k$, $k = 1,\ldots,K$, is a piecewise linear polynomial continuous at each knot.

- We can represent this model as
$$
y_i = \beta_0 + \beta_1 b_1(x_i) + \beta_2 b_2(x_i) + \cdots + \beta_{K+1} b_{K+1}(x_i) + \epsilon_i,
$$
where $b_k$ are **basis functions**:  
\begin{eqnarray*}
b_1(x_i) &=& x_i \\
b_{k+1}(x_i) &=& (x_i - \xi_k)_+, \quad k=1,\ldots,K.
\end{eqnarray*}
Here $(\cdot)_k$ means positive part
$$
(x_i - \xi_i)_+ = \begin{cases}
x_i - \xi_k & \text{if } x_i > \xi_k \\
0 & \text{otherwise}
\end{cases}.
$$

### Cubic splines

- A **cubic spline** with knots at $\xi_k$, $k = 1,\ldots,K$, is a piecewise cubic polynomial with continuous derivatives up to order 2 at each knot.

- Again we can represent this model with **truncated power basis functions**
$$
y_i = \beta_0 + \beta_1 b_1(x_i) + \beta_2 b_2(x_i) + \cdots + \beta_{K+3} b_{K+3}(x_i) + \epsilon_i,
$$
with
\begin{eqnarray*}
b_1(x_i) &=& x_i \\
b_2(x_i) &=& x_i^2 \\
b_3(x_i) &=& x_i^3 \\
b_{k+3}(x_i) &=& (x_i - \xi_k)_+^3, \quad k = 1,\ldots,K,
\end{eqnarray*}
where
$$
(x_i - \xi_i)_+^3 = \begin{cases}
(x_i - \xi_k)^3 & \text{if } x_i > \xi_k \\
0 & \text{otherwise}
\end{cases}.
$$

- A cubic spline with $K$ knots costs $K+4$ parameters or degrees of freedom. That is $4(K+1)$ polynomial coefficients minus $3K$ constraints.

- While the truncated power basis is conceptually simple, it is not too attractive numerically: powers of large numbers can lead to severe rounding problems. In practice, **B-spline basis functions** are preferred for their computational efficiency. See ESL Chapter 5 Appendix.



### Natural cubic splines

- Splines can have high variance at the outer range of the
predictors.

- A **natural cubic spline** extrapolates linearly beyond the boundary knots. This adds $4 = 2 \times 2$ extra constraints, and allows us to put more internal knots for the same degrees of freedom as a regular cubic spline.

- A natural spline with $K$ knots has $K$ degrees of freedom.

::: {.panel-tabset}



```{r}
#| code-fold: true
library(splines)

# Plot wage ~ age
Wage %>%
  ggplot(mapping = aes(x = age, y = wage)) + 
  geom_point(alpha = 0.25) + 
  # Cubic spline
  geom_smooth(
    method = "lm",
    formula = y ~ bs(x, knots = c(25, 40, 60)),
    color = "blue"
    ) +
  # Natural cubic spline
  geom_smooth(
    method = "lm",
    formula = y ~ ns(x, knots = c(25, 40, 60)),
    color = "red"
    ) +  
  labs(
    title = "Natural cubic spline fit (red) vs cubic spline fit (blue)",
    x = "Age",
    y = "Wage (k$)"
    )
```

:::

### Knot placement

- One strategy is to decide $K$, the number of knots, and then place them at appropriate quantiles of the observed $X$.

- In practice users often specify the degree of freedom and let software choose the number of knots and locations.

## Smoothing splines

- Consider this criterion for fitting a smooth function $g(x)$ to some data:
$$
\text{minimize} \quad \sum_{i=1}^n (y_i - g(x_i))^2 + \lambda \int g''(t)^2 \, dt.
$$
    - The first term is RSS, and tries to make $g(x)$ match the data at each $x_i$.  
    - The second term is a **roughness penalty** and controls how wiggly $g(x)$ is. It is modulated by the tuning parameters $\lambda \ge 0$. 
        * The smaller $\lambda$, the more wiggly the function, eventually interpolating $y_i$ when $\lambda = 0$.  
        * As $\lambda \to \infty$, the function $g(x)$ becomes linear. 
        
- The solution is a (shrunken) natural cubic spline, with a knot at every unique value of $x_i$. The roughness penalty still controls the roughness via $\lambda$. 

- Smoothing splines avoid the knot-selection issue, leaving a single $\lambda$ to be chosen.  

- The vector of $n$ fitted values can be written as $\hat{g}_\lambda = S_\lambda y$, where $S_{\lambda}$ is an $n \times n$ matrix (determined by the $x_i$ and $\lambda$). 

- The **effective degrees of freedom** are given by
$$
\text{df}_{\lambda} = \sum_{i=1}^n S_{\lambda,ii}.
$$
Thus we can specify `df` rather than $\lambda$.

- The leave-one-out (LOO) cross-validated error is given by
$$
\text{RSS}_{\text{CV}}(\lambda) = \sum_{i=1}^n \left[ \frac{y_i - \hat{g}_\lambda(x_i)}{1 - S_{\lambda,ii}} \right]^2.
$$


`ggformula` package supplies `geom_spline` function for displaying smoothing spline fits.
```{r} 
#| code-fold: true

library(ggformula)
library(splines)

# Plot wage ~ age
Wage %>%
  ggplot(mapping = aes(x = age, y = wage)) + 
  geom_point(alpha = 0.25) + 
  # Smoothing spline with df = 16
  geom_spline(
      df = 16,
      color = "red"
    ) +
  # Smoothing spline with GCV tuned df
  geom_spline(
    # df = 6.8,
    cv = TRUE,
    color = "blue"
    ) +
  labs(
    title = "Smoothing spline with df=16 (red) vs LOOCV tuned df=6.8 (blue)",
    x = "Age",
    y = "Wage (k$)"
    )
```


## Local regression

<p align="center">
![](ISL_fig_7_9.pdf){width=600px height=600px}
</p>

- With a sliding weight function, we fit separate linear fits over the range of $X$ by weighted least squares. 

- At $X=x_0$,
$$
\text{minimize} \quad \sum_{i=1}^n K(x_i, x_0) (y_i - \beta_0 - \beta_1 x_i)^2,
$$
where $K$ is a weighting function that assigns heavier weight for $x_i$ close to $x_0$ and zero weight for points furthest from $x_0$.

- **Locally weighted linear regression**: `loess` function in R and `lowess` in Python.

- Anecdotally, loess gives better appearance, but is $O(N^2)$ in memory, so does not work for larger data sets.

## Generalized additive model (GAM)

- Generalized additive models (GAMs) allows for flexible nonlinearities in several variables, but retains the additive structure of linear models.
$$
y_i = \beta_0 + f_1(x_{i1}) + f_2(x_{i2}) + \cdots + f_p (x_{ip}) + \epsilon_i.
$$

- We can fit GAM simply using, e.g. natural splines.

- Coefficients not that interesting; fitted functions are.

- Can mix terms: some linear, some nonlinear, and use ANOVA to compare models.

- Can use smoothing splines or local regression as well. In R:
`gam(wage ~ s(year; df = 5) + lo(age; span = :5) + education)`.

- GAMs are additive, although low-order interactions can be included in a natural way using, e.g. bivariate smoothers or interactions of the form (in R) `ns(age, df = 5):ns(year, df = 5)`.



Natural splines for `year` and `age`.
```{r}
gam_mod <- lm(
  wage ~ ns(year, df = 4) + ns(age, df = 5) + education,
  data = Wage
  )
summary(gam_mod)
```

Smoothing splines for `year` and `age`.
```{r}
library(gam)

gam_mod <- gam(
  wage ~ s(year, 4) + s(age, 5) + education,
  data = Wage
  )
summary(gam_mod)
plot(gam_mod, se = TRUE, col = "red")
```

